How to make an abridged life table and calculate life expectancy

### 2016-09-20 working_teaching_examples/how_to_abridged_life_table.Rmd
fixed up latex equations

$R_x=d_x/(l_x-d_x)+\bar{a_x}d_x$, dividing the numerator and denominator by l_x:

$R_x=q_x/(1-q_x)+\bar{a_x}q_x$, rearranging to solve for q_x:

@@

Observed mortality rates $R_x$ can therefore be used to calculate the probability of dying within each age-specific interval. The observed mortality rates are calculated as $d_x/Popn_x$, where $Popn_x$ is the number of individuals alive during age x. While this looks like a proportion, it calculates a rate when we assume the population is stationary, thereby implying there are $Popn_x$ person-years at risk of death during the age bracket.

@@

Recall that an abridged life table includes age intervals that are longer than a year. In practice, these intervals are often 5 years in length and can be represented by $n_x$, where $x$ is the age at the beginning of the interval. The formula for $q_x$ can be generalized to this context, where: $q_x=\frac{n_xR_x}{1+(1-\bar(a_x))n_xR_x}$.

@@

We can then substitute this expression for $l_x$ into the equations for $q_x$:

@@

```{r}
#load("Data/alabama_only.Rdata")
load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
#example data
alabama.by.strata <- 
  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
  arrange(RaceSex, Year2, Age2) %>%
  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years")) %>%
  summarise(total_deaths5 = sum(Count_md5),
            Population = first(Population), 
            Race2 = first(Race2), 
            Sex2 = first(Sex2), 
            StateYearRaceSex = first(StateYearRaceSex),
            nx = first(nx))

>>>>

$R_x=d_x/(l_x-d_x)+\bar{a_x}d_x$, dividing the numerator and denominator by $l_x$:

$R_x=q_x/(1-q_x)+\bar{a_x}q_x$, rearranging to solve for $q_x$:

@@

Observed mortality rates $R_x$ can therefore be used to calculate the probability of dying within each age-specific interval. The observed mortality rates are calculated as $d_x/Pop_x$, where $Pop_x$ is the number of individuals alive during age $x$. While this looks like a proportion, it calculates a *rate* when we assume the population is stationary, thereby implying there are $Pop_x$ person-years at risk of death during the age bracket.

@@

Recall that an abridged life table includes age intervals that are longer than a year. In practice, these intervals are often 5 years in length and can be represented by $n_x$, where $x$ is the age at the beginning of the interval. The formula for $q_x$ can be generalized to this context, where: $q_x=\frac{n_xR_x}{1+(1-\bar{a_x})n_xR_x}$.

@@

We can then substitute this expression for $l_x$ into the equation for $q_x$:

$q_x=\frac{d_x}{l_x}$

@@

```{r alabama_example}
#load("Data/alabama_only.Rdata")
load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
#example data
alabama.by.strata <- 
  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
  arrange(RaceSex, Year2, Age2) %>%
  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years")) %>%
  summarise(total_deaths5 = sum(Count_md5),
            Population = first(Population), 
            Race2 = first(Race2), 
            Sex2 = first(Sex2), 
            StateYearRaceSex = first(StateYearRaceSex),
            nx = first(nx))
